\documentclass{article}
\usepackage{amsmath}

\title{Structural Type Inference: If It Quacks Like a Duck...}
\author{Jay R Bolton}

\addtolength{\oddsidemargin}{-.875in}
\addtolength{\evensidemargin}{-.875in}
\addtolength{\textwidth}{1.75in}
\addtolength{\topmargin}{-.875in}
\addtolength{\textheight}{1.75in}

\begin{document}
\maketitle

\section*{Introduction}

Python is a wildly dynamic language whose program behavior depends almost
entirely on the execution path. Despite being very multi-paradigm, it is also
not a very complicated language, with an abstract grammar of only around 100
lines. The enforcement of type correctness in Python is performed at run-time
using Duck Typing, where we are only concerned with the comparative structure
of two objects. We'll explore this further in sections below.

All languages can be statically typed, no matter how dynamic their current
implementation may be. Our goal here is to perform static type checking (albeit
in a limited form) to a very dynamic language with very minimal inherent type
enforcement. Our type checking system will be optional and will not affect the
ability to run the program. The advantages of having such a system are:

\begin{itemize}
\item Retain `agile development' without limiting possible correct programs.
\item Still be able to analyze the type correctness of your programs, catching errors that you may not find in tests.
\item Provide documentation and meaningful type annotations that can be helpful
in understanding your program in a number of different ways, discussed in the
last section.
\item Potentially increase the performance of the language using the static type information.
\end{itemize}

This project is not concerned specifically with any one of those items. It
serves more as a demonstration of pure attribute type inference.

\subsection*{Terms and Definitions}

When I use the term \emph{static}, I am referring to something that
\textbf{does not depend} upon the execution of the program. When I say
\emph{dynamic}, I mean something that depends upon or arises from the execution
of the program.

A \emph{type} is a labeled set of data bound by certain properties. In this
project, those properties will be object attributes, which we'll discuss in
more detail below.

\section*{The Type System of Python}

Python is a 100\% dynamic language and as such uses no static type information.
The interpreter parses and then executes your code, and any errors arise as
they are encountered.

Everything in Python is an object, which provides a nice uniformity for our
type system.

\subsection*{Duck Typing}

Duck Typing, or structural typing, is simply types represented as sets of attributes.

An attribute is an object-oriented term that refers to the data fields in a class or object. Thus a type is a set of all the data fields for some object. For example:

\begin{verbatim}
class A(object):
  x = 1
	y = 'a'
\end{verbatim}

The class A is given two fields, x and y, so it has the type: $\{x : int, y : str\}$, which is a two-element set of keys and values, where the keys are \emph{attribute labels} and the values are \emph{other types}.

Duck typing is usually defined as a ``dynamic typing'' system, but this does
not necessarily have to be true.  The language Scala allows programmers to
write explicit, static duck types as parameters to functions:

\begin{verbatim}
def f(obj: { def x; def y }) = {
	  obj.x
		obj.y
}
\end{verbatim}

This is a Scala function that takes an object with two fields: x and y. We are not concerned with any other properties of `obj' except that it has those two fields. We do not even care what its superclass might be.

\subsection*{Type Enforcement}

In Python, type errors occur almost entirely when we reference attributes for some object. For example:

\begin{verbatim}
a = A()
a.x # int
a.q # undefined
\end{verbatim}

We reference two attributes of the object `a.' First, we reference x, which is
defined and has type `int.' We then reference the undefined attribute `q.' This
will return a type error when the Python interpreter tries to execute it.

Undefined references comprise the vast majority of type errors in python. Our
goal will then be to infer these sets of attributes for all the objects in a
program and do lookups on references to make sure they exist.

\section*{Type Inference Using Only Attributes}

Structural type inference is somewhat of a conceptual branching off from
traditional Hindley-Milner type inference found in Haskell and ML. In
describing the rules for structural type inference, I will try to use similar
notation as found in the papers on Hindley-Milner, but will have to alter much
of it to fit this different paradigm.

\subsection*{The Grammar}

Our type grammar is fairly simple; we use sets of pairs to represent types, and
map those to labels.

The `type environment,' or the set of all types inferred so far, is simply a
type itself. It can also be said to be the type of the module, or the file, of
the program.

\begin{align*}
& Type = \alpha\{ Name : Type, \dots, Name : Type \} \\
& Name = [a-z] \\
& Environment = Type \\
& List = [Type, Type, \dots, Type] \\
& Tuple = [Type, Type, \dots, Type] \\
& Error = ``message"
\end{align*}

The $\alpha$ represents a type label and may be any lower-case Greek letter.

We must distinguish between \emph{type labels} and \emph{attribute names}.  An
attribute name is paired with both a type label and a type. 
We may have two attributes with two different names but with the same type. We
may also have two attributes with the same name but with different types. It is
therefore necessary to be able to distinguish types in and of themselves.

Certain built-in Python types---Lists, Tuples, and Dictionaries---are special cases. They
are container objects whose items we would like to be able to see all at once
in the type signature, so we store a list of the types of its contents.

Recall that in Python, everything is an object, including functions and
classes. A function type is not a special case in this system; it is simply an
object type with attributes that make it callable. A class object is also
callable, and the return value is the type of its instances.

\subsection*{Notation}

We'll use the capital gamma, or $\Gamma$, to denote our `environment,' which
includes all the built-ins, such as types for addition and subtraction.

We'll use capital Greek letters to indicate whole types, lower case Greek
letters for type labels, and Latin letters to indicate attribute names.

We use a crossbeam symbol, $\Gamma \vdash\ e\ :\ T$, to indicate that we are able
to derive the type $T$ for attribute $e$ from the environment $\Gamma$.

A long right arrow indicates the application of a rule. Above the arrow, we
have the assumptions and conditions; below it is the conclusion that we can draw using 
those assumptions.

For example:

\begin{align*}
\Gamma\ \vdash\ a:T, T \vdash\ b:E\\
\Longrightarrow \\
c:E \vdash \Gamma
\end{align*}

States that if we assume we can derive the type of attribute $a$ from the
environment $\Gamma$ to be of type $T$, and the type of attribute $b$ from $T$ to be $E$, then we can say that attribute $c$ has type $E$, and that we can insert that mapping back into $\Gamma$.

The $\vdash$ operator is overloader for our purposes. If the left hand side is
a type, then we are doing a lookup on the attributes in that type. If the left
hand side is a pair of attribute mapped to type, then we are inserting that
pair into the type on the right hand side.

Assumptions above the arrow are separated by commas or line breaks and often
start with the actual syntax we are defining a rule for.

\subsection*{Lookup and Assignment}

We'll start with the most basic case: the `identity', or simple look-up of a
single object, which is a sort of axiom or first principle rather than a rule.

\begin{align*}
&Lookup\\
& \Gamma \vdash\ e\ :\ T
\end{align*}

The assignment rule is the next simplest. The type of the left hand side of an
assignment is derived from the type of the right hand side.

\begin{align*}
&Assignment\\
& \texttt{x = e},\ \Gamma\ \vdash\ e:T\\
& \Longrightarrow \\
& x\ :\ T \vdash \Gamma \\
\end{align*}

Attribute reference is a look-up of the type of the object. We first get the
type of the object `x' from the environment, and then get the type of the
attribute from the type of x.

\begin{align*}
&Attribute\\
& \texttt{x.y},\ \Gamma\ \vdash\ x:T,\ T \vdash y:\Upsilon\\
& \Longrightarrow \\
& \texttt{x.y}\ :\ \Upsilon \\
\end{align*}

The next step is function definition, which is also a large step forward in complexity.

In order to define this rule, we need to have a new notation for \emph{open
types}. Open types are types that will absorb new attributes when they
are referenced on them. In other words, when we reference an undefined
attribute on an object, that object then acquires that attribute name mapped to
an empty type inside the object's own type.

The notation we can use for an open type is $(T)$.

Here is the open type version of the attribute referencing rule:

\begin{align*}
&OpenAttribute\\
& \texttt{x.y},\ \Gamma\ \vdash\ x:(T),\ (T) \vdash y:None \\
& \Longrightarrow \\
& \texttt{x.y}\ :\ \alpha\{\},\ y\ :\ \alpha\{\} \vdash (T) \\
\end{align*}

Now we can define our large function definition inference rule:

Capital delta represents our ``scoped environment.'' It is a copy of the environment with the parameter types overwritten, or ``shadowed.''

\begin{align*}
&FunctionDefinition\\
& \texttt{def f(p):b},\ \Gamma = \Delta, \\
& p = (p_1,\dots,p_n, p) : Tuple([(\alpha_1\{\}),\dots,(\alpha_n\{\})]) \vdash \Delta \\
& \Delta \vdash b : T,\ T \vdash \texttt{return}:E \\
& T \vdash p : X \\
& \Longrightarrow \\
& f : \{*r : E, *p : X\} \vdash \Gamma
\end{align*}

First we take a duplicate environment $\Delta$, and insert a tuple type of
empty open types into it, representing our shadowed parameters.

We want our parameter types to be open so that any attribute references in the
body of the function modify their type.

`*r' and `*p' represent special meta-attributes for the return type and the
parameter type which we can use for callable objects.

Next up is the actual function call, which is when we apply our function
definition types. Luckily, it is slightly simpler.

\begin{align*}
&FunctionCall\\
& \texttt{f(a)},\ \Gamma \vdash f : T \\
& \Gamma \vdash a : X,\ T \vdash *p : X' \\
& X \approx X' = S \\
& S \rightarrow T \vdash \texttt{return} : E\\
& \Longrightarrow \\
& \texttt{f(a)} : E
\end{align*}

This rule introduces a couple new symbols. $\approx$ is used to denote
\emph{unification}, which is explained in the next section. In the third line
we assume that $X$ unifies with $X'$. The right arrow denotes that the
unification produces a substitution $S$. In the forth line, we apply the
substitution $S$ to the function type $T$, and then derive from that our return
type.

Our final type inference rule is that of class definition.

\begin{align*}
&ClassDefinition\\
& \texttt{class C: b},\ \Gamma = \Delta, \\
& \Delta \vdash b : T,\ T \vdash \texttt{init}:E \\
& T \vdash \texttt{self} : X \\
& \Longrightarrow \\
& C : T + \{*r : X + E \}
\end{align*}

$X + E$ represents the merging of types X and E, where duplicate types in E overwrite those in X.

In this rule, we derive the type of our body to be $T$, and then look up the
type of the constructor or \texttt{init} function from within the type of the
body. For the the attributes of the instance objects, we look up the type of
`self.' We combine the type from the constructor with the one from self to put
into the return field for the class. We give the class itself all the remaining
fields.

This last rule is quite rough and does not contain all the details in the
implementation of class definition type inference. For example, in order for
the open type `self' to propogate up into the type of the class body, we would
have to add a special case in function definition. Also in python, assignments
in the body provide fields for both the class and the instances.  Despite its
lack of accuracy, hopefully the rule will convery the general idea.

\subsection*{Unification and Substitution}

Unification is used in the function call rule to check that the given
function type and the applied type are compatible.

Unification produces a substitution, which is a mapping of labels to types,
indicating that the types for each of the lables should be replaced with those
listed.

\begin{align*}
Substitution = \{a_1 \mapsto s_1, \dots, a_n \mapsto s_n\}
\end{align*}

Unification itself is performed according to the following rules:

\begin{enumerate}
\item An attribute `a' with an empty type $A$ unifies with any other type $T$ , producing $\{a \mapsto T\}$ 
\item When unifying non-empty types $a : \Upsilon$ and $T$, all attributes in $T$ must be
present and unifiable in $\Upsilon$. This also produces the substitution $\{a \mapsto T\}$
\item List and tuple types must unify their contained types element-wise. For example: the list $l : \Lambda$ unified with $T$ produces $\{l \mapsto T\}$ where all types inside $\Lambda$ and $T$ are unifiable element-wise.
\end{enumerate}

A unification of types $X$ and $Y$ produce a substitution $S$, denoted $X
\approx Y = S$. We can apply that substitution to a type $Y$ with $S
\rightarrow Y = Y'$, where $Y'$ is the type $Y$ but with all substitutions in
$S$ applied.

\section*{Examples and Cases}

Our example will use the shortest code possible that demonstrates all the type inference rules. We'll walk through the heuristics required to then infer all the types in the environment (also known as the type of the module).

\begin{minipage}{\linewidth}
\begin{verbatim}
class C:
  x = 1
  __init__(self):
    self.y = "s"
    self.z = 2
  smeth(x,y): return x
  imeth(self): return self.x
c = C()
d = C.smeth(9,"q")
e = c.imeth()
f = C.none()
\end{verbatim}
\end{minipage}

We'll briefly walk through the application of the type inference rules. In
order to save space, we won't step through the actual rules, we'll just show
their conclusions.  The algorithm will go top-down, and upon reaching names and
primitives, will then traverse bottom-up.

\begin{align*}
Assignment,FunctionDefinition,ClassDefinition \\
\Longrightarrow\\
\Gamma \vdash C : \alpha\{x:1,\ *r:\{y:str, z:int, x:int\}\, *p:Tuple[]\}
\end{align*}

Analysis of the body of the class produces a type for $C$ that is callable.
This callable type has an empty parameter tuple and a return type that
represents the type of the instances for the class.

\begin{align*}
FunctionCall, Assignment \Longrightarrow \Gamma \vdash c : \alpha\{y:str, z:int, x:int\}
\end{align*}

Our constructor produces the instance type of class C. 

\begin{align*}
FunctionCall, Assignment \Longrightarrow \Gamma \vdash d : int
\end{align*}

Calling the static method \texttt{smeth} returns the type of the first parameter, an integer.

\begin{align*}
FunctionCall, Assignment \Longrightarrow \Gamma \vdash  e : int
\end{align*}

Calling the instance method \texttt{imeth} returns the type of the attribute `x' inside `c', with is an integer. Note that in the function call, `c' is bound to `self' in the first parameter.

\begin{align*}
FunctionCall, Assignment \Longrightarrow f : Error\ "undefined"
\end{align*}

When we try to reference an undefined attribute ``none'' on `c', then we return an Error type with a message.

\section*{Improvement and Expansion}

The two major features not included in this system are rules for module
importing and rules for super classes.


\section*{The Implementation}

This system has been implemented in Jython (a Python interpreter that produces
byte code for the Java Virtual machine), located at
\texttt{https://github.com/aesacus/python-type-inferencer}. The design of that program is described in a presentation at:

\texttt{https://docs.google.com/presentation/d/19vKKkBxZPOV1TP4DecFaXtIoMewjXlAVXo2GhZ0t0j0}.

The above program works well on a pre-written demo of sample Python code. It
could use expansion and further testing, as well as more readable and usable
data output. It comes with a demo script (\texttt{demo.py}) that showcases its
features.

\section*{References and Further Reading}

I derived most of the rules and functions myself. The basic structure of type
inference, where you pass around an environment, apply unification and get
substitutions, was learned from Cardelli.

\begin{itemize}
\item Luca Cardelli. ``Basic Polymorphic Typechecking''
\item Bryan Hackett and Shu-yu Guo. ``Fast and Precise Hybrid Type Inference for Javascript''
\end{itemize}

\end{document}
